---
title:
author:
date:
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, message=FALSE, results='hide'}
library(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
source("r/show-solution.R")
# ATTENZIONE: leggere nota dentro show-solution.R
```


## Data Object

### Vectors
#### Exercise 1

a. Create a vector, named _vec1_, containing the following values: 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90.  
b. Select the 5-th element of _vec1_
c. Select the first 10 elements of _vec1_
d. Select all the elements of _vec1_ apart from the 2nd and the 6th elements

```{r ex1-vec, echo=TRUE, message=FALSE, results='hide'}
vec1 <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90)
vec1[5]
vec1[1:10]
vec1[-c(2,6)]
```

#### Exercise 2

a. Generate a vector, named _vec2_, containing numbers from 1 to 10 of length 8, using the operator `seq`.
b. Select the values of _vec2_ which are greater than 4
c. Select the values of _vec2_ which are equal or less than 2 or the values that are equal or greater than 6

```{r ex2-vec, echo=TRUE, message=FALSE, results='hide'}
vec2 <- seq(from=1, to=10, length.out = 8)
vec2[vec2>4] # or y > 4 b[y]
vec2[vec2<=2 | vec2>=6]
```

#### Exercise 3

a. Generate the following vector using the operator `rep`: `vec3 <- c("one", "two", "one", "two", "one", "two")`
b. Generate a new vector, named _vec5_, combining the previous vector, _vec3_, with the following one: `vec4 <- c("three", "four")`

```{r ex3-vec, echo=TRUE, message=FALSE, results='hide'}
vec3 <- rep(c("one", "two"), times=3)
vec4 <- c("three", "four")
vec5 <- c(vec3, vec4)
vec5
```

### Matrices

#### Exercise 1

Generate a matrix, named _mat1_, with 5 rows and 3 columns, using `matrix` function

```{r ex1a-mat}
mat1 <- matrix(1:15, nrow = 5, ncol = 3, byrow = TRUE)
mat1
```


```{r ex1b-mat, echo=show_solution, eval=show_solution}
mat1 <- matrix(1:15, nrow = 5, ncol = 3, byrow = TRUE)
```


#### Exercise 2

Starting from the following vector: `mat2 <- 1:8`, generate a matrix with 2 rows and 4 columns using the function `dim` 

```{r ex2-mat, echo=show_solution, eval=show_solution}
mat2 <- 1:8
dim(mat2) <- c(2,4)
mat2
```


#### Exercise 3

a. Generate a matrix, named _mat3_, combining the following columns:
```{r ex3a-mat}
a <- 1:3
b <- 7:9
c <- 8:6
```

```{r ex3b-mat, echo=show_solution, eval=show_solution}
mat3 <-cbind(a,b,c)
```

b. Add the following row to _mat3_: `d <- 4:6`


```{r ex3c-mat, echo=show_solution, eval=show_solution}
mat3 <-rbind(mat3, d)
```

#### Exercise 4

Considering the following matrix, named _mat4_:

```{r ex4a-mat}
mat4 <- matrix(1:24, nrow = 6, ncol = 4, byrow = TRUE)
mat4
```

a. Select the second and the sixth row of _mat4_  
b. Select all columns of _mat4_ apart from the first 
c. Select second and third rows and second and third columns of _mat4_

```{r ex4b-mat, echo=show_solution, eval=show_solution}
mat4[c(2,6),]
mat4[, -1]
mat4[2:3, 2:3] # or mat4[c(2,3) , c(2,3)] 
```

### Lists

#### Exercise 1

a. Generate a list, maed _list1_ the contains the following R elements:

```{r ex1a-list}
vec <- 1:10
mat <- matrix(1:9, ncol = 3)
name <- "Oscar"
```

b. Add the element `letters <- c("a", "b", "c", "d")` to _list1_

```{r ex1b-list, echo=show_solution, eval=show_solution}
list1 <- list(vec = 1:10, mat = matrix(1:9, ncol = 3), name = "Oscar")
list1
```

### Exercise 2

Given the following list, named _list2_:
```{r ex2a-list, echo=show_solution, eval=show_solution}
list2 <- list(vec = c(1,3,5,7,8), mat = matrix(1:12, ncol = 4), sub_list = list(names = c("Veronica", "Enrico", "Andrea", "Anna"), numbers = 1:4))
list2
```

a. Entract the first element of _list2_
b. Extract the objects contained in the third element of _list2_
c. Extract the object named _sub\_list_ from _list2_
d. Extract the second rows of the matrix included in the second element of _list2_


```{r ex2b-list, echo=show_solution, eval=show_solution}
list2[2]
list2[[3]]
list2$sublist
list2[[2]][2,] # or list2$mat[2,]
```


### Factors

#### Exercise 1

Starting from the vector `fac1 <- c("F", "F", "M", "M" , "F")`, generate the corresponding factor with two levels: "F" and "M"

```{r ex1a-factors, echo=show_solution, eval=show_solution}
fac1 <- c("F", "F", "M", "M" , "F")
fac1 <- factor(fac1, levels = c("F", "M"))
fac1
```

### Exercise 2

a. Starting from the vector `fac2 <- c(1, 1, 1, 2, 2, 2)`, generate the corresponding factor considering that 1 = "Female", 2 = "Male" e 3 = "Trans"
b. Select the all elements of the factor apart from "Male" 
```{r ex6a-list, echo=show_solution, eval=show_solution}
fac2 <- c(1, 1, 1, 2, 2, 2) 
fac2 <- factor(fac2, levels = c(1,2,3), labels = c("Female", "Male", "Trans"))
fac2
fac2[fac2!= "Male"]
```



```{r ex6b-list, echo=show_solution, eval=show_solution}
```



