---
title: 
author:
date: 
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
show_solution <- FALSE
# ATTENZIONE: leggere nota dentro show-solution.R
```
 
# Data Object
## Vectors
### Exercise 1

a. Create a vector, named `vec1`, containing the following values:  
1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90    
```{r ex1a-vec, echo=show_solution, eval=show_solution}
vec1 <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90)
```

b. Select the 5-th element of `vec1`. 
```{r ex1b-vec, echo=show_solution, eval=show_solution}
vec1[5]
```

c. Select the first 10 elements of `vec1.`  
```{r ex1c-vec, echo=show_solution, eval=show_solution}
vec1[1:10]
```

d. Select all the elements of `vec1` apart from the 2nd and the 6th element.  

```{r ex1d-vec, echo=show_solution, eval=show_solution}
vec1[-c(2,6)]
```

### Exercise 2

a. Generate a vector, named `vec2`, containing the numbers from 1 to 10 and of length 8, using the function `seq()`.  
```{r ex2a-vec, echo=show_solution, eval=show_solution}
vec2 <- seq(from=1, to=10, length.out = 8)
```

b. Select the values of `vec2` which are greater than 4.  
```{r ex2b-vec, echo=show_solution, eval=show_solution}
vec2[vec2>4] # or y > 4; b[y]
```

c. Select the values of `vec2` which are equal or less than 2 or which are equal or greater than 6.  

```{r ex2c-vec, echo=show_solution, eval=show_solution}
vec2[vec2<=2 | vec2>=6]
```

### Exercise 3

a. Generate the following vector using the function `rep()`:   
`vec3 <- c("one", "two", "one", "two", "one", "two")`  

```{r ex3a-vec, echo=show_solution, eval=show_solution}
vec3 <- rep(c("one", "two"), times=3)
```

b. Generate a new vector, named `vec5`, combining the previous vector, `vec3`, with the following one:   
 
```{r ex3b1-vec}
vec4 <- c("three", "four")
```

```{r ex3b2-vec, echo=show_solution, eval=show_solution}
vec5 <- c(vec3, vec4)
vec5
```

## Matrices

### Exercise 1

Generate a matrix, named `mat1`, with 5 rows and 3 columns, using `matrix()` function:

```{r ex1a-mat, echo=FALSE}
mat1 <- matrix(1:15, nrow = 5, ncol = 3, byrow = TRUE)
mat1
```


```{r ex1b-mat, echo=show_solution, eval=show_solution}
mat1 <- matrix(1:15, nrow = 5, ncol = 3, byrow = TRUE)
```


### Exercise 2

Starting from the following vector:  
```{r ex2a-mat}
mat2 <- 1:8
```

Generate a matrix with 2 rows and 4 columns using `dim()` function. 

```{r ex2b-mat, echo=show_solution, eval=show_solution}
dim(mat2) <- c(2,4)
mat2
```


### Exercise 3

a. Generate a matrix, named `mat3`, combining the following columns:
```{r ex3a-mat}
a <- 1:3
b <- 7:9
c <- 8:6
```

```{r ex3b-mat, echo=show_solution, eval=show_solution}
mat3 <- cbind(a,b,c)
mat3
```

b. Add the following row to `mat3`: 

```{r ex3c1-mat}
d <- 4:6
```


```{r ex3c2-mat, echo=show_solution, eval=show_solution}
mat3 <-rbind(mat3, d)
mat3
```

### Exercise 4

Considering the following matrix, named `mat4`:

```{r ex4a-mat}
mat4 <- matrix(1:24, nrow = 6, ncol = 4, byrow = TRUE)
mat4
```

a. Select the third and the fifth row of `mat4`.  
```{r ex4b1-mat, echo=show_solution, eval=show_solution}
mat4[c(3,5),]
```

b. Select all columns of `mat4` apart from the first.  
```{r ex4b2-mat, echo=show_solution, eval=show_solution}
mat4[, -1]
```

c. Select second and third rows and second and third columns of `mat4`.  

```{r ex4b3-mat, echo=show_solution, eval=show_solution}
mat4[2:3, 2:3] # or mat4[c(2,3) , c(2,3)] 
```

## Lists

### Exercise 1

a. Generate a list, named `list1` that contains the following R elements:

```{r ex1a1-list}
vec <- 1:10
mat <- matrix(1:9, ncol = 3)
name <- "Oscar"
```

```{r ex1a2-list, echo=show_solution, eval=show_solution}
list1 <- list(vec = 1:10, mat = matrix(1:9, ncol = 3), name = "Oscar")
list1
```

b. Add to `list1` the following element:

```{r ex1b1-list}
letters <- c("a", "b", "c", "d")
```

```{r ex1b2-list, echo=show_solution, eval=show_solution}
list1$letters <- letters
list1
```


### Exercise 2

Given the following list, named `list2`:
```{r ex2a-list}
list2 <- list(vec = c(1,3,5,7,8), mat = matrix(1:12, ncol = 4), 
              sub_list = list(names = c("Veronica", "Enrico", "Andrea", "Anna"), 
                              numbers = 1:4))
list2
```

a. Entract the first element of `list2`.  
```{r ex2b1-list, echo=show_solution, eval=show_solution}
list2[1]
```

b. Extract the objects contained in the first element of `list2`.  
```{r ex2b2-list, echo=show_solution, eval=show_solution}
list2[[1]]
```

c. Extract the element named `sub_list` of `list2`.  
```{r ex2b3-list, echo=show_solution, eval=show_solution}
list2$sub_list
```

d. Extract the second rows of the matrix included in the second element of `list2`.  
```{r ex2b4-list, echo=show_solution, eval=show_solution}
list2[[2]][2,] # or list2$mat[2,]
```


## Factors

### Exercise 1

Starting from the vector:  
```{r ex1a-factors}
fac1 <- c("F", "F", "M", "M" , "F")
```

Generate the corresponding factor with two levels: "F" and "M"

```{r ex1-factors, echo=show_solution, eval=show_solution}
fac1 <- factor(fac1, levels = c("F", "M"))
fac1
```

### Exercise 2

Starting from the vector:  
```{r ex2a1-factors}
fac2 <- c(1, 1, 1, 2, 2, 2) 
```

a. Generate the corresponding factor considering that 1 = "Female", 2 = "Male" e 3 = "Trans".  
```{r ex2a2-factors, echo=show_solution, eval=show_solution}
fac2 <- factor(fac2, levels = c(1,2,3), labels = c("Female", "Male", "Trans"))
fac2
```

b. Select the all elements of `fac2` apart from "Male".     
```{r ex2b-factors, echo=show_solution, eval=show_solution}
fac2[fac2!= "Male"]
```

## Data Frames


### Exercise 1

a. Generate a data frame, named `df1`, corresponding to:

```{r ex1a1-df, echo=FALSE}
df1 <- data.frame(id=1:10, 
                  name=c("Luca", "Chiara", "Lisa", "Matteo", "Alice", "Marco", 
                         "Veronica", "Nicola", "Elena", "Daniele"),
                  class=c(rep("5A", times=5), rep("5B", times=5)),
                  mean= c(6,7,5,6.5,7.5,4.5, 9, 8, 8.5, 7), stringsAsFactors = FALSE)
df1
```

Remember to maintain character vectors as they are, specifiyng `stringsAsFactors = FALSE`.    
```{r ex1a2-df, echo=show_solution, eval=show_solution}
df1 <- data.frame(id=1:10, 
                  name=c("Luca", "Chiara", "Lisa", "Matteo", "Alice", "Marco", 
                         "Veronica", "Nicola", "Elena", "Daniele"),
                  class=c(rep("5A", times=5), rep("5B", times=5)),
                  mean= c(6,7,5,6.5,7.5,4.5, 9, 8, 8.5, 7), stringsAsFactors = FALSE)
df1

# Other solution
id <- 1:10
name <- c("Luca", "Chiara", "Lisa", "Matteo", "Alice", "Marco", 
          "Veronica", "Nicola", "Elena", "Daniele")
class <- c(rep("5A", times=5), rep("5B", times=5))
mean <- c(6,7,5,6.5,7.5,4.5, 9, 8, 8.5, 7)
df1 <- data.frame(id, name, class, mean, stringsAsFactors = FALSE)
df1
```


b. Select the first 3 rows of `df1`.  
```{r ex1b-df, echo=show_solution, eval=show_solution}
df1[1:3,]
```

c. Select the last 6 rows and the first 3 columns of `df1`.    
```{r ex1c-df, echo=show_solution, eval=show_solution}
df1[5:10, 1:3]
```

d. Select the column `class` of `df1`.   
```{r ex1d-df, echo=show_solution, eval=show_solution}
df1$class
```

e. Convert the column `class` of `df1` in a factor with levels: "5A" and "5B"
```{r ex1e-df, echo=show_solution, eval=show_solution}
df1$class <- factor(df1$class, levels = c("5A", "5B"))
df1$class
```

f. How many columns and rows `df1` has?   
```{r ex1f-df, echo=show_solution, eval=show_solution}
dim(df1) # or ncol(df1) and nrow(df1)
```

g. Generate another dataframe, named `df2` composed by the columns `name` and `mean` of `df1`, specifying the argument `stringsAsFactors = FALSE`.   
```{r ex1g-df, echo=show_solution, eval=show_solution}
df2 <- data.frame(name = df1$name, mean=df1$mean, stringsAsFactors = FALSE)
df2
```

h. Show the first rows and the structure of `df2`.    
```{r ex1h-df, echo=show_solution, eval=show_solution}
head(df2)
str(df2)
```





