---
title: "Data Programming Exam"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
show_solution <- TRUE
```

Please reply to the following questions in an R script called "surname_name.R" (e.g., Mario Rossi
will return a file named "rossi\_mario.R") and send it by mail to andrea.spano@quantide.com    

Please comment your answers using the symbol: #, before the comments (e.g., # my comment).  

## Exercise 1

a. Create a vector, named `vec`, containing the following values:  
1, 5, 12, 14, 6, 78, 68, 34, 34, 32, 56, 75

```{r ex1_a, eval=show_solution, echo=show_solution}
vec <- c(1, 5, 12, 14, 6, 78, 68, 34, 34, 32, 56, 75)
```

b. Select the 3-rd element of `vec`.

```{r ex1_b, eval=show_solution, echo=show_solution}
vec[3]
```

c. Select all elements of `vec` apart from the 1st.

```{r ex1_c, eval=show_solution, echo=show_solution}
vec[-1]
```

## Exercise 2

a. Generate a matrix, named `mat`, with 3 rows and 5 columns containig numbers from 1 to 15. 

```{r ex2_a, eval=show_solution, echo=show_solution}
mat <- matrix(1:15, nrow = 3, ncol = 5)
```

b. Select 2-nd and 3-rd rows and 1-st and 3-rd columns of `mat`.

```{r ex2_b, eval=show_solution, echo=show_solution}
mat[c(2,3), c(1,3)]
```

## Exercise 3

Given the following list, named `this_list`:

```{r ex3_question1}
this_list <- list(numbers = c(2,3,5,6,7), letters = c("z", "x", "y", "t"))
this_list
```

a. Extract the element named `letters` of `this_list` by using the `$` operator.

```{r ex3_a, eval=show_solution, echo=show_solution}
this_list$letters
```

b. Extract the first element of `this_list` by using double square brackets.

```{r ex3_b, eval=show_solution, echo=show_solution}
this_list[[1]]
```


\clearpage

## Exercise 4

a. Generate a data frame, named `df`, corresponding to:

```{r ex4_a, echo=show_solution, comment=''}
df <- data.frame(country = c("Italy", "France", "China", "Japan", "Libya", 
                              "Cameroon"),
                 population = c(59801004, 64668129, 1382323332, 126323715,
                                 6330159, 23924407),
                 continent = c("Europe", "Europe", "Asia", "Asia", 
                                "Africa", "Africa"
                               ), stringsAsFactors = FALSE)
print(df, row.names = FALSE)
```

  Use `data.frame()` function and remember to maintain character vectors as they are, specifiyng `stringsAsFactors = FALSE`.

b. Convert `continent` variable of `df` as a factor with levels: "Europe", "Asia" and "Africa". Use `factor()` function.

```{r ex4_b, eval=show_solution, echo=show_solution}
df$continent <- factor(df$continent, levels = c("Europe", "Asia", "Africa"))
```


## Exercise 5

a.  Import the file _2008.csv_ into a data frame named `flights` by using the `read.table()` command. Remember to specify `stringsAsFactors` as `FALSE` in order to avoid importing character columns as factors.  
  Before importing be sure about:
    - column names in the first row
    - the field separator
    - the decimal separator

```{r ex5_a, eval=show_solution, echo=show_solution}
flights <- read.table(file = "/home/veronica/dev/qtraining/010-rbase/data/2008.csv", header = TRUE, sep = ",", quote = "", dec = ".")
```

This dataset contains information about flight arrival and departure details for all commercial flights within the USA in 2008.

Load `dplyr` library:

```{r ex5_question, message=FALSE}
require(dplyr)
```


b. Convert `flights` data frame to a `tbl_df` using `tbl_df()` function.

```{r ex5_b, eval=show_solution, echo=show_solution}
flights <- flights %>% tbl_df()
```

c. Starting from `flights` data frame, select `ArrDelay` and `Dest` variables and filter the records for which `ArrDelay` variable is greater than 120.

```{r ex5_c, eval=show_solution, echo=show_solution}
flights %>% select(ArrDelay, Dest) %>% filter(ArrDelay > 120)
```

d. Starting from `flights` data frame, compute the mean delay at departure (`DepDelay` variable) grouping by `Origin` variable. Remember to add `na.rm=TRUE option` to mean computation.

```{r ex5_d, eval=show_solution, echo=show_solution}
flights %>% group_by(Origin) %>% summarise(mean=mean(DepDelay, na.rm=TRUE))
```

\clearpage

## Exercise 6

Load `mtcars` dataset in this way:

```{r ex6_question1}
data("mtcars")
```

`mtcars` data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

```{r ex6_question2}
head(mtcars)
```

To achieve more information about `mtcars` dataset type `?mtcars` on R console.

Load `ggplot2` library:

```{r ex6_question3, message=FALSE}
require(ggplot2)
```

a. Calculate the number of rows and columns of the `mtcars` dataset.

```{r ex6_a, eval=show_solution, echo=show_solution}
dim(mtcars) # or nrow(mtcars) and ncol(mtcars)
```

b. Build a scatterplot to analyze the relationship between `mpg` and `wt` variables. Use `ggplot()` and `geom_point()` functions.

```{r ex6_b, eval=show_solution, echo=show_solution}
ggp <- ggplot(data = mtcars, mapping = aes(x=wt, y=mpg)) + 
  geom_point()
print(ggp)
```

c. Represent the distribution of `mpg` variable with an histogram. Use `ggplot()` and `geom_histogram()` functions.

```{r ex6_c, eval=show_solution, echo=show_solution}
ggp <- ggplot(data = mtcars, mapping = aes(x=mpg)) + 
  geom_histogram()
print(ggp)
```








