---
title: "A First Look to R Session"
output: html_document
---

```{r options, include=FALSE, purl=FALSE}
source("options.R")
```


## The First R Session

Start the R system, the cursor is waiting for you to type in some R commands. For example, use R as a simple calculator:

```{r, calculator}
6 + 3
5 - 9
4 * 6
8 / 3
5 ^ 2
(1 + 0.05)^8
exp(3)
log(14)
23.76 * log(8)/(23 + atan(9))
```




## Assignment

Results of calculations can be stored in objects using the assignment operators:
  
 - an arrow (`<-`) formed by a smaller than character and a hyphen without a space;
 - the equal character (`=`).

These objects can then be used in other calculations. 

There are some restrictions when giving an object a name:
  
 - Object names cannot contain "strange" symbols like `!`, `+`, `-`, `#`.
 - A dot (`.`) and an underscore (`_`) are allowed, also a name starting with a dot.
 - Object names can contain a number but cannot start with a number.
 - R is case sensitive: `X` and `x` are two different objects, as well as `temp` and `temP`.

```{r, assignment}
x <- log(14)
y <- 23.76 * log(8)/(23 + atan(9))
z <- x + y
```
To print the object just enter the name of the object or use `print()` function. 

```{r, print_obj}
x
y
print(z)
```

## The R Workspace

Objects created during an R session are hold in memory. The collection of objects currently avalaible is called the workspace.
To list the objects in the current R session, the function `ls()` or the function `objects()` may be used.

```{r, workspace}
ls()
objects()
```

So to run the function `ls` the name followed by an opening and a closing bracket must be entered. Entering only `ls` will just print the object, i.e. the underlying R code of the function.

Most functions in R accept certain arguments. For example, one of the arguments of the function `ls()` is pattern. To list all objects starting with the letter `x`:
  
```{r, ls}
x2 <- log(9)
ls(pattern = "x")
```


If a value to an object that already exists is assigned then the contents of the object will be overwritten with the new value (without a warning!). The function `rm()` ought be used to remove one or more objects from your session.

```{r, rm}
rm(x2)
ls()
```


## The Working Directory
<!-- questo va cambiato AS -->

If you want to read files from a specific location or write files to a specific location you will need to set working directory in R. The following example shows how to set the working directory in R to the folder "Data" within the folder "Documents and Settings" on the C drive.

```{r, setwd, eval=FALSE}
setwd("C:/Documents and Settings/Data")
```

To get the path of the current working directory, the `getwd()` function should be used.

```{r, getwd, eval=FALSE}
getwd()
```


```
## [1] "C:/Documents and Settings/Data"
```

Remember that you must use the forward slash / or double backslash \\\\ in R! The Windows format of single backslash will not work.


## R help

Within R, the following functions provide help about R itself:

 - The HTML version of R's online documentation can be printed on-screen by typing `help.start()`;
 - Online documentation for most of the functions and variables in R can be printed on-screen by typing `help(name)` (or `?name`), where _name_ is the name of the topic help is sought for;
 - Online documentation for finding help pages on a vague topic can be printed on-screen by typing `help.search('topic')`;
 - A list of function containing _topic_ in the name can be printed on-screen by typing `apropos('topic')`;
 - A research in the website can be performed by typing `RSiteSearch('query')`, where _query_ is the search query.

To get help about the `mean()` function, the `help()` function can be used.

```{r, help1, eval=FALSE}
help(mean)
```

The `help()` function can be called using the `?`.
```{r, help2, eval=FALSE}
?mean
```
To get a list of functions concerning the mean, the `help.search()` function can be used.

```{r, help.search1, eval=FALSE}
help.search("mean")
```

To get a list of function containing "mean" in the name, the `apropos()` function can be used.

```{r, help.search2, eval=FALSE}
apropos("mean")
```

