---
title: "Function Object"
output: html_document
---

Functions are one of the most important objects in R as when working with R we all make constant use of functions.

A function is characterised by an input and an output. The input of the function, i.e. the set of the arguments, can be either null or made of one or more R objects. The output of the function can be either null or a single R object. If the function has to return more than one object, the objects are to be inserted in a list.

The function `ls()` doesn't require any input and returns an output.

```{r, argument1}
ls()
```

The function `rm()` requires an input and doesn't return an output.

```{r, argument2}
foo <- 3
rm(foo)
```

Usually, functions require an input and returns an output.

```{r, argument3}
sum(1:10)
```

The several elements of the input of a function are the arguments of the function. When a function is called the arguments can be declared explicitely or implicitely. The arguments can be explicitely with their name.

```{r, argument4}
sum(c(1, 2, NA), na.rm = TRUE)
```
In the above example, the `na.rm` argument of the `sum()` function is called explicitely.

When arguments are called implicitely, the order of the arguments is fundamental. The third example of the code below returns a wrong result.

```{r, argument5}
rdata <- c(rnorm(100), NA)
quantile(x = rdata, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)
quantile(rdata, c(0.25, 0.50, 0.75), TRUE)
quantile(rdata, TRUE, c(0.25, 0.50, 0.75))
# When arguments are called explicitely, you can change arguments order 
quantile(rdata, na.rm = TRUE, probs = c(0.25, 0.50, 0.75))
```


It is advisable to use the help command of each function to understand not only the arguments which can be used as function inputs, but also the output:
  
```{r, help3, eval=F}
help(read.table)
```

or

```{r, help4, eval=F}
?read.table
```

There lots of function types, let us see some examples.

### Mathematical Functions

The names of the basic functions and mathematical operators in R follow the standards of programming languages. In this paragraph the functions and operators enabling basic mathematical operations to be performed will be dealt with. R can also be used to perform more complex calculations, such as matrix operations or calculations with complex numbers.

Functions are usually applied to one or more vectors. In this case, operations are performed on each element of each vector. Vectors have to be the same length.

```{r, operations}
x <- 1:10
y <- 11:20
z <- -4:5
x + y + z
exp(x)
log(z)
abs(z)
sqrt(x)
```

The `sum()` function calculates the sum of all the elements of a vector.

```{r, sum}
sum(x)
```

The `floor`, `ceiling`, `trunc` and `round` functions can be used to round a number. `floor()` returns a numeric vector containing the largest integers not greater than the corresponding input elements. `ceiling()` returns a numeric vector containing the smallest integers not less than the corresponding input elements. `trunc()` returns a numeric vector containing the integers formed by truncating the input values toward 0. `round()` rounds the values in its first argument to the specified number of decimal places (default 0).

```{r, round}
floor(3.14)
floor(3.67)
ceiling(3.14)
ceiling(3.67)
trunc(3.14)
trunc(3.67)
round(3.14, digits = 1)
round(3.19, digits = 1)
```


### Statistical Functions

Any kind of statistical analysis can be performed in R thanks to the built-in functions of the base version or the numerous additional packages. The functions enabling the calculation of the main descriptive statistical analyses are explained below.
 
The `mean()`, `median()`, `sd()` and `var()` functions are used to calculate the mean, the median, the sample standard deviation and the sample variance of a numeric vector. 

```{r, stats}
x <- mtcars$mpg
mean(x)
median(x)
sd(x)  
var(x)
```

The `quantile()` function calculates one or more quantiles.

```{r, quantile}
quantile(x, .9) 
quantile(x, c(.3, .84))
quantile(x, c(.25, .50, .75))
```

The `min()` and `max()` functions return the minimum and maximum  value respectively.

```{r, minmax}
min(x)
max(x)
```

The `summary()` generic function applied to a numeric vector returns minimum, maximum, quartiles and arithmetic mean.

```{r, summary}
summary(x)
```

Correlation and covariance can be calculated with the `cor()` and `cov()` functions, respectively.

```{r, corcov}
data <- mtcars[, c(1, 3, 4, 5, 6)]
cor(data)
cov(data)
```



