---
title: "R Data Files"
---

```{r, options, echo=FALSE, results='hide', message=FALSE, warning=FALSE, purl=FALSE}
options(width = 108)
require(knitr)
opts_knit$set(root.dir = "./../data")
outDir <- "."
```
 

## Save R Data Files 

Statistical packages often provide the opportunity to save the working environment with all the objects it contains in their own formats. Even if rarely used, this function is available in R as well. The format used by R is called `Rdata` (or `Rda`).

In this way, different objects can be saved in a single file. Moreover, all the features of a data frame which cannot be saved in a text file, such as the levels of a factor, can be kept in the file. 

To save an object of the R workspace in a file use the `save()` function. The first argument of the function is the object to be saved, whereas the file name is defined in the `file` argument. If the position is not specified, R saves the file in the current directory.

```{r, save1, eval=FALSE}
# It creates a mtcars.Rda file in the current directory
save(mtcars, file = "mtcars.Rda")
```

To save more than one object list their names.

```{r, save2, eval=FALSE}
# It creates a datasets.Rda file in the current directory
save(mtcars, iris, file = "datasets.Rda")
```

An alternative method to save more than one object is provided by the `list` argument. The names of the objects to be saved in a vector can be inserted with the `list` argument. This method is advisable when the list of the files to be saved is contained in a vector.

```{r, save3, eval=FALSE}
# It creates a datasets.Rda file in the current directory
datalist = c("mtcars", "iris")
save(list = datalist, file = "datasets.Rda")
```

## Load R Data Files

To upload `Rda` files in R use the `load()` function.

```{r, load, eval=FALSE}
# It reads the datasets.Rda file previously created in the current directory
load("datasets.Rda")
```

