---
title: "Chaining Verbs"
---

```{r options, include=FALSE, purl=FALSE}
source("options.R")
```

```{r first, include=FALSE, purl=TRUE, message=FALSE}
# This code chunk contains R code already described in the previous chapters
# that is required by following examples
require(dplyr)
require(qdata)
data(bank)
```


# Chain Together Multiple Operations: the `%>%` Operator

The `dplyr` API is functional in the sense that function calls don’t have side-effects, and you must always save their results. This doesn’t lead to particularly elegant code if you want to do many operations at once. You either have to do it step-by-step:

```{r}
a1 <- group_by(bank, date)
a2 <- select(a1, age, balance)
a3 <- summarise(a2,
                mean_age = mean(age, na.rm = TRUE),
                mean_balance = mean(balance, na.rm = TRUE)
                )
(a4 <- filter(a3, mean_age < 40 & mean_balance > 5000))
```

Or if you don’t want to save the intermediate results, you need to wrap the function calls inside each other:

```{r}
filter(
  summarise(
    select(
      group_by(bank, date), age, balance
    ),
    mean_age = mean(age, na.rm = TRUE),
    mean_balance = mean(balance, na.rm = TRUE)
  ),
  mean_age < 40 & mean_balance > 5000
)
```

This is difficult to read because the order of the operations is from inside to out, and the arguments are a long way away from the function. To get around this problem, `dplyr` provides the `%>%` operator. `x %>% f(y)` turns into `f(x, y)` so you can use it to rewrite multiple operations so you can read from left-to-right, top-to-bottom:

```{r}
bank %>%
  group_by(date) %>%
  select(age, balance) %>%
  summarise(
    mean_age = mean(age, na.rm = TRUE),
    mean_balance = mean(balance, na.rm = TRUE)
  ) %>%
  filter(mean_age < 40 & mean_balance > 5000)
```
