---
title: "Writing R Functions"
---

## Writing a function

A function in R can be defined with the `function()` function. Its use can be summed up as follows:
  
```{r functions-writing2}
myFunction =  function(a, b, c) {...}
```

* `myFunction` is the name of the customised function 
* `a`, `b` and `c` are the arguments of `myFunction()` 
* the part in curly brackets contains the definition of the function

As an example, write a function which calculates taxable income and taxes, given the total and the VAT percentage.

```{r functions-writing-example, tidy=FALSE}
vat = function(amount, rate = 0.21) {
  taxable = amount / (1 + rate)
  tax = amount - taxable
  return(list(tax = tax, taxable = taxable))
}
vat(121)
vat(104, rate = 0.04)
```

The value assigned to the `rate` argument is the default value assumed by the argument itself.

It is good practice to end the function specifying its output with the `return()` or `invisible()` functions. The function output ought to be made of a single object. If there are multiple objects, they can be inserted in a list. 







