
```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
show_solution <- FALSE
# ATTENZIONE: leggere nota dentro show-solution.R
```
 
# Your first R session 

## Aritmetic with R

### Exercise 1

Calculate your body mass index dividing your body mass (kg) by the square of your body height (m) ($kg/m^2$)

```{r ex1a-ar, echo=show_solution, eval=show_solution}
65/1.7^2
```

## Assignment

### Exercise 1

a. Assign your age (in number) to `age` variable.

```{r ex1a-ass, echo=show_solution, eval=show_solution}
age <- 26
```

c. Print out the value of the variable `age`.

```{r ex1b-ass, echo=show_solution, eval=show_solution}
age # o print(age)
```

d. Remove the variable `age` from the workspace, by using `rm()` function.

```{r ex1c-ass, echo=show_solution, eval=show_solution}
rm(age)
```

### Exercise 2

Suppose you want to buy 10 roses and 8 sunflowers in a flower shop. The roses cost 3 euros each and the sunflowers 2 euros each.  

a. Assign the total cost of roses to `roses_cost` variable and the total cost of sunflowers to `sunflowers_cost` variable. 

```{r ex2a-ass, echo=show_solution, eval=show_solution}
roses_cost <- 10 * 3
sunflowers_cost <- 8 * 2
```

b. Calculate the total cost of flowers by adding `roses_cost` and `sunflowers_cost` variables and assign it to `flowers_cost` variable.

```{r ex2b-ass, echo=show_solution, eval=show_solution}
flowers_cost <- roses_cost + sunflowers_cost
```

c. Print out the value of the variable `flowers_cost`.

```{r ex2c-ass, echo=show_solution, eval=show_solution}
flowers_cost
```

d. List the objects in the current R session, by using `ls()` function.

```{r ex2d-ass, echo=show_solution, eval=show_solution}
ls() # or objects()
```







