---
title: 
author:
date: 
output:
  html_document:
    self_contained: no
---


```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(tidyverse) 
require(knitr) 
opts_knit$set(message=FALSE)
opts_knit$set(warning=FALSE)
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
opts_knit$set(root.dir="./data")
show_solution <- if (exists("show_solution_mk")){
  show_solution_mk
} else {
  FALSE
}


# ATTENZIONE: leggere nota dentro show-solution.R
```



Combining data
================================

## Joins: `inner_join()`, `left_join()`, `right_join()`, etc.
Note: all the exercises of this section are based on `flights`, `airlines`, `airports` or `planes` datasets.


```{r, message = FALSE}
library(dplyr)
library(nycflights13)
```


### Exercise 1
Keep only the following variables of the `flights` dataset: month, day, hour, origin, destination and carrier. Save this dataset in a data frame and call it `flights_red`. Through a proper join command, add the carrier name to `flights_red` (this piece of information is available in `airlines`).


```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_red <- flights %>% 
  select(month, day, hour, origin, dest, carrier)

right_join(flights_red, airlines)                      
```



### Exercise 2
Through a proper join command, add name, latitude, longitude and altitude of the origin airport to `flights_red` (these pieces of information are available in `airports`). Do the same also for the destination airport. (If you are able to, try to keep variables about both origin and destination airports in the same final dataset).


```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_red %>% left_join(airports, c("origin" = "faa"))                     
```

```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_red %>% left_join(airports, c("dest" = "faa"))                      
```


### Exercise 3
Through the `inner_join()` function, redo the same for the destination airport but keep only the flights whose information is available in both datasets (`flights` and `airports`).


```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_red %>% inner_join(airports, c("dest" = "faa"))                     
```


### Exercise 4
Redo the exercise 3 by using `full_join()` instead of `inner_join()`. What is the difference in the result?

```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_red %>% full_join(airports, c("dest" = "faa"))

# there are a few more rows due to the fact that full_join keeps all 
# rows even those with no matches
```



### Exercise 5
Through the `anti_join()` function, extract all the flights from `flights` whose information about destination airport is not available in `airports`.


```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_red %>% anti_join(airports, c("dest" = "faa"))
```



### Exercise 6
Sort the `planes` dataset by increasing year. Then create two datasets: the first will deal with planes older than 2000; the second will deal with planes of 2000 or newer. Finally create a unique dataset where the first rows will deal with the newest planes, whereas the last rows will deal with the oldest planes.

```{r, echo = show_solution, eval = show_solution, cache=FALSE}

planes_old <- planes %>% 
  arrange(year) %>%
  filter(year <= 2000)

planes_young <- planes %>% 
  arrange(year) %>%
  filter(year > 2000)

planes_old %>% bind_rows(planes_young)

```


