---
title:
author:
date:
output:
  html_document:
    self_contained: no
---


```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(tidyverse) 
require(knitr) 
options(width=80)
opts_knit$set(message=FALSE)
opts_knit$set(warning=FALSE)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
opts_knit$set(root.dir="./data")
show_solution <- if (exists("show_solution_mk")){
  show_solution_mk
} else {
  FALSE
}


# ATTENZIONE: leggere nota dentro show-solution.R
```

Dates with lubridate
================================

## lubridate
Note: all the exercises of this section are based on the `flights` dataset.


```{r, message = FALSE}
require(tidyverse)
require(lubridate)
require(nycflights13)
```


### Exercise 1

Using the flights data, build the variable `dep_date` based on the variables `year`, `month` and `day`. First use the function `unite()` and then the  parsing function `ydm()`. Select only the new variable and save the new data frame called `flights_date`.


```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_date <- flights %>% 
    unite(date, year, month, day) %>% 
    mutate(date = ymd(date)) %>%
    select(date)  
```



### Exercise 2
Using the dataset, shift all flights by two months. Save it in a separate data frame called `flights_date_2`.

```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_date_2 <- flights_date %>% mutate(date2 = date + months(2)) 
```



### Exercise 3
Take the new date (2 months ahead) and substruct the orginal variable `date`. `flights_date_2`.


```{r, echo = show_solution, eval = show_solution, cache=FALSE}
flights_date_2 %>% mutate(date2-date)
```


