---
title:
author:
date:
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, message=FALSE, results='hide'}
library(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
source("r/show-solution.R")
# ATTENZIONE: leggere nota dentro show-solution.R
```


## `select()` and its friends
Note: all the exercises of this section are based on the `flights` dataset.

```{r require, echo=TRUE, message=FALSE, results='hide'}
library(dplyr)
library(nycflights13)
```


### Exercise 1
Extract the following information about flights:

* month;
* day;
* air_time;
* distance.

```{r ex1, echo=show_solution, eval=show_solution}
select(flights, month, day, air_time, distance)

# Alternative solution
vars <- c("month", "day", "air_time", "distance")
select(flights, one_of(vars))
```


### Exercise 2
Extract all information about flights except hour and minute.

```{r ex2, echo=show_solution, eval=show_solution}
select(flights, -hour, -minute)

# Alternative solution
vars <- c("hour", "minute")
select(flights, -one_of(vars))
```


### Exercise 3
Select all variables whose name ends in "time".

```{r ex3, echo=show_solution, eval=show_solution}
select(flights, ends_with("time"))
```


### Exercise 4
Select all variables whose name contains the word “delay”.

```{r ex4, echo=show_solution, eval=show_solution}
select(flights, contains("delay"))
```


### Exercise 5
Select the `tailnum` variable and rename it into `tail_num`.  

```{r ex5, echo=show_solution, eval=show_solution}
select(flights, tail_num = tailnum)
```


### Exercise 6
Select all the variables and rename the `tailnum` variable into `tail_num`.  

```{r ex6, echo=show_solution, eval=show_solution}
rename(flights, tail_num = tailnum)
```

