---
title:
author:
date:
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(tidyverse) 
require(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
opts_knit$set(root.dir="./data")
show_solution <- if (exists("show_solution_mk")){
  show_solution_mk
} else {
  FALSE
}


# ATTENZIONE: leggere nota dentro show-solution.R
```



## `select()` and its friends
Note: all the exercises of this section are based on the `flights` dataset.


```{r}
require(tidyverse)
require(nycflights13)
```



### Exercise 1
Extract the following information about flights:

* month;
* day;
* air_time;
* distance.


```{r, echo = show_solution, eval = show_solution}
# require(nycflights13)
data(flights)

flights %>% select(month, day, air_time, distance)
```




### Exercise 2
Extract all information about flights except hour and minute.

```{r, echo = show_solution, eval = show_solution}
flights %>% select(-hour, -minute)
```



### Exercise 3
Select all variables whose name ends in "time".


```{r, echo = show_solution, eval = show_solution}
flights %>% select(ends_with("time"))
```



### Exercise 4
Select all variables whose name contains the word “delay”.


```{r, echo = show_solution, eval = show_solution}
flights %>% select(matches("delay"))
```



### Exercise 5
Select the `tailnum` variable and rename it into `tail_num`.  

```{r, echo = show_solution, eval = show_solution}
flights %>% select(tail_num = tailnum)
```



### Exercise 6
Select all the variables and rename the `tailnum` variable into `tail_num`.  
```{r, echo = show_solution, eval = show_solution}
flights %>% rename(tail_num = tailnum)
```


