---
title:
author:
date:
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
source("r/show-solution.R")
# ATTENZIONE: leggere nota dentro show-solution.R
```

## `filter()` and its friends
Note: all the exercises of this section are based on the `flights` dataset.

```{r require, results='hide', message=FALSE}
library(dplyr)
library(nycflights13)
```


### Exercise 1
Select all flights which delayed more than 1000 minutes at departure.

```{r ex1, echo=show_solution, eval=show_solution}
filter(flights, dep_delay > 1000)
```


### Exercise 2
Select all flights which delayed more than 1000 minutes at departure or at arrival.

```{r ex2, echo=show_solution, eval=show_solution}
filter(flights, dep_delay > 900 | arr_delay > 900)
```


### Exercise 3
Select all flights which took off from "EWR" and landed in "IAH" on Christmas Day.

```{r ex3, echo=show_solution, eval=show_solution}
filter(flights, month == 12 & day == 25 & origin == "EWR" & dest == "IAH")
```


### Exercise 4
Select the first five flights in this dataset.

```{r ex4, echo=show_solution, eval=show_solution}
slice(flights, 1:5)
```


### Exercise 5
Select the last ten flights in this dataset.

```{r ex5, echo=show_solution, eval=show_solution}
slice(flights, (n()-9):n())
```


### Exercise 6
Extract information about distance for all flights which delayed more than 1000 minutes at departure.

```{r ex6, echo=show_solution, eval=show_solution}
select(filter(flights, dep_delay > 1000), distance)
```

