---
title:
author:
date:
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
source("r/show-solution.R")
# ATTENZIONE: leggere nota dentro show-solution.R
```

## `mutate()` and its friends
Note: all the exercises of this section are based on the `flights` dataset.  
Times are in minutes and distances are in miles.

```{r require, results='hide', message=FALSE}
library(dplyr)
library(nycflights13)
```


### Exercise 1
Add the following new variables to the `flights` dataset:

* the gained time in minutes, defined as the difference between delay at departure and delay at arrival;
* the speed in miles per hour (`distance` / `air_time` * 60).

Show only the following variables: delay at departure, delay at arrival, distance, air time and the two new variables (gained time and speed).

```{r ex1, echo=show_solution, eval=show_solution}
flights2 <- mutate(flights,
                   gained_time = dep_delay - arr_delay,
                   speed = distance / air_time * 60)
select(flights2, dep_delay, arr_delay, gained_time, distance, air_time, speed)
```


### Exercise 2
Redo the previous calculations keeping only the new variables.

```{r ex2, echo=show_solution, eval=show_solution}
transmute(flights,
          gained_time = dep_delay - arr_delay,
          speed = distance / air_time * 60)
```


### Exercise 3
After sorting flights in chronological order, for each flight calculate the difference between its delay at arrival and the delay at arrival of the immediately previous flight. Have R showed only the delay variables (delay at departure, delay at arrival and the new variable).

```{r ex3, echo=show_solution, eval=show_solution}
sorted_flights <- arrange(flights, year, month, day, hour, minute)
delay <- select(sorted_flights, dep_delay, arr_delay)
mutate(delay, delta_delay = dep_delay - lag(dep_delay))
```


### Exercise 4
For each flight calculate the 'min ranking' in terms of delay at arrival.

```{r ex4, echo=show_solution, eval=show_solution}
mutate(delay, delay_rank = min_rank(arr_delay))
```


### Exercise 5
For each flight calculate the 'first ranking' in terms of delay at arrival.

```{r ex5, echo=show_solution, eval=show_solution}
mutate(delay, delay_rank = row_number(arr_delay))
```


### Exercise 6
Create a variable which indicates if each flight took off more than -4 or less than 4 minutes late.

```{r ex6, echo=show_solution, eval=show_solution}
mutate(delay, delay_rank = between(dep_delay, -3, 3))
```
