---
title: 
author:
date: 
output:
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, results='hide', message=FALSE}
library(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))
source("r/show-solution.R")
# ATTENZIONE: leggere nota dentro show-solution.R
```

Chainining verbs
================================

## Chain together multiple operations (`%>%`)
Note: all the exercises of this section are based on the `flights` or the `weather` datasets.

```{r require, results='hide', message=FALSE}
library(dplyr)
library(nycflights13)
```


### Exercise 1
Calculate number of flights, mean flights distance, mean delay at departure and at arrival for each of the three origin airports.

```{r ex1, echo=show_solution, eval=show_solution}
flights %>%
  group_by(origin) %>%
  summarise(n = n(),
            mean_distance = mean(distance),
            mean_dep_delay = mean(dep_delay, na.rm = TRUE),
            mean_arr_delay = mean(arr_delay, na.rm = TRUE))
```


### Exercise 2
Calculate number of flights and monthly mean gained time in minutes, where the gained time is defined as the difference between delay at departure and delay at arrival.

```{r ex2, echo=show_solution, eval=show_solution}
flights %>%
  group_by(year, month) %>%
  summarise(n = n(),
            mean_dep_delay = mean(dep_delay, na.rm = TRUE),
            mean_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  mutate(mean_gained_time = mean_dep_delay - mean_arr_delay)
```


### Exercise 3
Select all days where the mean delay at departure or the mean delay at arrival is greater than 30 minutes.

```{r ex3, echo=show_solution, eval=show_solution}
flights %>%
  group_by(year, month, day) %>%
  select(arr_delay, dep_delay) %>%
  summarise(
    arr = mean(arr_delay, na.rm = TRUE),
    dep = mean(dep_delay, na.rm = TRUE)
    ) %>%
  filter(arr > 30 | dep > 30)
```


### Exercise 4
From the `weather` dataset select the days taken off from EWR in 2013 where minimum visibility was 5 or lower.

```{r ex4, echo=show_solution, eval=show_solution}
weather %>%
  filter(origin == "EWR") %>%
  group_by(month, day) %>%
  summarise(min_visib = min(visib, na.rm = TRUE)) %>%
  filter(min_visib <= 5)
```

