---
title: "Introduction to Shiny"
output: html_document
---

```{r options, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'H', fig.width = 4, fig.height = 3)
options(width = 108)
```


```{r first, include=TRUE, purl=TRUE, message=FALSE}
require(tidyverse)
```


## Introduction


Shiny is a package from RStudio that allows you to easily build interactive web pages with R. You can easily interact and display R objects: plots, tables, etc. To get a glimpse of the many things you can build with Shiny, out of the many websites that contain Shiny examples, we suggest you visit https://daattali.com/shiny/ and https://shiny.rstudio.com/gallery/.


If you want to find additional material, you may visit the RStudio official Shiny page https://shiny.rstudio.com/ and we suggest you always keep the Shiny  http://shiny.rstudio.com/images/shiny-cheatsheet.pdf cheat sheet on hand when you are developing your Shiny app!

In this second part of the course, we'll walk you through the creation of a Shiny app using a dataset on comics characters. We will start from a basic app and we'll add complexity step by step. Of course you may think of adding additional features as we go on and as you think of additional data insights!


## Required R packages

Before we begin you need to install `Shiny` package:

```{r, eval = FALSE}
install.packages("shiny")
```

To ensure you successfully installed Shiny, try running one of the demo apps.

```{r, eval = FALSE}
library(shiny)
runExample("01_hello")
```

If the example app is running, press Escape to close the app, and you are ready to build your first Shiny app!



## Comic characters data

The data we will visualise in the Shiny App are the raw data behind the story [Comic Books Are Still Made By Men, For Men And About Men](https://fivethirtyeight.com/features/women-in-comic-books/).

Originally, the data come from Marvel Wikia and DC Wikia, characters were scraped on August 24, appearance counts were scraped on September 2 and the month and year of the first issue each character appeared in was pulled on October 6.

The data is split into two files although here, for cimplicity we will consider only one file, `marvel-wikia-data.csv`.

The dataset `marvel-wikia-data.csv` is contained in the data folder of this manual. Using `readr` package, load the dataset as this is the data we'll work on to build our shiny app:


```{r load-data}
require(tidyverse)
df <- read_csv("../data/marvel-wikia-data.csv")
```


Let us quickly explore the dataset so that you can start thinking of intersting data visualisations:

```{r data-explore}
str(df)
```


The data comes from Marvel Wikia. The file was scraped in August 2014 and contains the following variables:

* `page_id`: The unique identifier for that characters page within the wikia
* `name`:	The name of the character
* `urlslug`:	The unique url within the wikia that takes you to the character
* `id`:	The identity status of the character (Secret Identity, Public identity, [on marvel only: No Dual Identity])
* `align`:	If the character is Good, Bad or Neutral
* `eye`:	Eye color of the character
* `hair`:	Hair color of the character
* `sex`:	Sex of the character (e.g. Male, Female, etc.)
* `gsm`:	If the character is a gender or sexual minority (e.g. Homosexual characters, bisexual characters)
* `alive`:	If the character is alive or deceased
* `appearances`:	The number of appareances of the character in comic books (as of Sep. 2, 2014. Number will become increasingly out of date as time goes on.)
* `first-appearances`	The month and year of the character's first appearance in a comic book, if available
* `year`:	The year of the character's first appearance in a comic book, if available
