---
title: "Introduction to Shiny"
output: html_document
---

```{r options, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'H', fig.width = 4, fig.height = 3)
options(width = 108)
```


```{r first, include=TRUE, purl=TRUE, message=FALSE}
require(tidyverse)
```


## Introduction


Shiny is a package from RStudio that allows you to easily build interactive web pages with R. You can easily interact and display R objects: plots, tables, etc. To get a glimpse of the many things you can build with Shiny, out of the many websites that contain Shiny examples, we suggest you visit https://daattali.com/shiny/ and https://shiny.rstudio.com/gallery/.


If you want to find additional material, you may visit the RStudio official Shiny page https://shiny.rstudio.com/ and we suggest you always keep the Shiny app https://www.rstudio.com/resources/cheatsheets/ cheat sheet on hand when you are developing your Shiny app!

In this second part of the course, we'll walk you through the creation of a Shiny app using a dataset on comics characters. We will start from a basic app and we'll add complexity step by step. Of course you may think of adding additional features as we go on and as you think of additional data insights!


## Required R packages

Before we begin you need to install `Shiny` package:

```{r, eval = FALSE}
install.packages("shiny")
```

To ensure you successfully installed Shiny, try running one of the demo apps.

```{r, eval = FALSE}
library(shiny)
runExample("01_hello")
```

If the example app is running, press Escape to close the app, and you are ready to build your first Shiny app!



## Data

The data folder contains the dataset `marvel-wikia-data.csv` that is a collection of . Using `readr` package, load the dataset as this is the data we'll work on to build our shiny app:


```{r load-data}
require(tidyverse)
df <- read_csv("../data/marvel-wikia-data.csv")
```


Let us quickly explore the dataset so that you can start thinking of intersting data visualisations:

```{r data-explore}
str(df)
```


The data comes from Marvel Wikia. The file was scraped in August 2014 and contains the following variables:

* `page_id`: The unique identifier for that characters page within the wikia
* `name`:	The name of the character
* `urlslug`:	The unique url within the wikia that takes you to the character
* `ID`:	The identity status of the character (Secret Identity, Public identity, [on marvel only: No Dual Identity])
* `ALIGN`:	If the character is Good, Bad or Neutral
* `EYE`:	Eye color of the character
* `HAIR`:	Hair color of the character
* `SEX`:	Sex of the character (e.g. Male, Female, etc.)
* `GSM`:	If the character is a gender or sexual minority (e.g. Homosexual characters, bisexual characters)
* `ALIVE`:	If the character is alive or deceased
* `APPEARANCES`:	The number of appareances of the character in comic books (as of Sep. 2, 2014. Number will become increasingly out of date as time goes on.)
* `FIRST APPEARANCE`	The month and year of the character's first appearance in a comic book, if available
* `YEAR`:	The year of the character's first appearance in a comic book, if available
