---
title: "Introduction to Graphics"
output: html_document
---

```{r options, include=FALSE, purl=FALSE}
source("options.R")
```

```{r first, include=FALSE, purl=TRUE}
# This code chunk contains R code already described in the previous chapters
# that is required by following examples

require(dplyr)
```

R provides at least three main graphical systems: `graphics` that comes with base R, `lattice` that is an R implementation of William Cleveland's trellis graphics and `ggplot2` that is an R implementation of Leland Wilkinson's Grammar of Graphics. The Grammar of Graphics is a formal and structured perspective on how to describe data graphics. CRAN provides a [Task View on Graphics](https://cran.r-project.org/web/views/Graphics.html), that is an organized list of all R packages about graphics, that includes not only the three main graphical systems but also packages to produce specialized plots.

In the following chapters, `ggplot2` graphics are shown. There are at least two main reasons to prefer `ggplot2` that other graphic systems:

 - `ggplot2` is newer than other graphic systems;
 - `ggplot2` is built on the idea of a semantics for graphics and there is much more emphasis on reshaping data, transformation, and assembling the elements of a plot.

`ggplot2` was written by Hadley Wickham and Winston Chang. If you are wondering, Wickham wrote also a `ggplot` package (without any number) that was archived by CRAN in 2008.

The starting point for more information about `ggplot2` is the [official website](http://ggplot2.org/) providing documentation, the official mailing list and other resources.

Supposing the package is already installed, first of all `ggplot2` must be loaded.  

```{r intro_require_ggplot, message=FALSE}
require(ggplot2)
```

Each of the following chapters introduces a different plot with one or more related geometric object (`geom`). Moreover, each chapter introduces some basic concept about `ggplot2`, according to the following table.

| Plot          | Geoms | `ggplot2` key concepts introduced in the chapter                                                    |  
|---------------|-------|-----------------------------------------------------------------------------------------------------|
| Scatter plots | `geom_point` `geom_smooth`                         | aesthetics, _setting_ or _mapping_, geoms, assignment  |
| Line plots   | `geom_line` `geom_point` `geom_hline` `geom_vline` | `group` aesthetic, manual scales                       |
| Bar plots    | `geom_bar` `geom_text`                             | `stat`, hide legend, swapping axes                     |
| Histograms    | `geom_histogram` `geom_density`                    | faceting                                               |
| Box plots     | `geom_boxplot`                                     | titles and theme (plot title and axes ticks/titles)    |

The last chapter of the section shows how to prepare plots for presentations, and save them to files.


# Datasets

The following data sets are used in the examples of this section.

## `bands`

`bands` provides data about process delays known as cylinder banding in rotogravure printing.

`bands` data is included in the `qdata` package.

```{r intro_data_bands, message=FALSE}
require(qdata)
data(bands)
bands
```

For a complete description of data, refer to the help (`?bands`).


## `ChickWeight`

`ChickWeight` provides weight versus age of chicks on different diets.

`ChickWeight` data is included in `datasets` package, that comes with base R: you just need to type `ChickWeight` to access data.

For a description of data, refer to the help (`?ChickWeight`).


## `ToothGrowth`

`ToothGrowth` provides the effect of Vitamin C on Tooth Growth in Guinea Pigs.

`ToothGrowth` data is included in `datasets` package, that comes with base R: you just need to type `ToothGrowth` to access data.

For a description of data, refer to the help (`?ToothGrowth`).



# Bibliography and References

To take a deeper look about `ggplot2` you can refer to:

 - [Chang W., _R Graphics Cookbook_, O'Reilly, 2012](http://www.cookbook-r.com/Graphs/) that provides a practical approach to `ggplot` and other R plots;
 - [Wickham H., _ggplot2: Elegant Graphics for Data Analysis_, Springer, 2009](http://ggplot2.org/book/) that provides technical details about Grammar of Graphics implemented in `ggplot2`.
 
