

```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=7, fig.height=7, fig.pos='H!'))#, fig.path="figures/lm-"))

show_solution <- show_solution <- if (exists("show_solution_mk")){
  show_solution_mk
} else {
  FALSE
}
```


# Data Objects

## Data Frames, Vectors and Factors

### Exercise 1 
a. Generate a data frame, named `df`, corresponding to:

```{r ex1a, echo=show_solution}
df <- data.frame(country = c("Italy", "France", "China", "Japan", "Libya", "Cameroon"),
                 population = c(59801004, 64668129, 1382323332, 126323715, 6330159, 23924407),
                 continent = c("Europe", "Europe", "Asia", "Asia", "Africa", "Africa"), 
                 stringsAsFactors = FALSE)
df
```

Remember to maintain character vectors as they are, specifiyng `stringsAsFactors = FALSE`.    
```{r ex1b, echo=show_solution, eval=show_solution}
df <- data.frame(country = c("Italy", "France", "China", "Japan", "Libya", "Cameroon"),
                 population = c(59801004, 64668129, 1382323332, 126323715, 6330159, 23924407),
                 continent = c("Europe", "Europe", "Asia", "Asia", "Africa", "Africa"), 
                 stringsAsFactors = FALSE)
df
```

b. Supposing `dplyr` package is already installed, convert the previously defined data frame in tbl_df class.

```{r ex1c, message=FALSE}
require(dplyr)
```

```{r ex1d, echo=show_solution, eval=show_solution}
df <- tbl_df(df)
```

### Exercise 2 

a. Generate a numeric vector, named `num_vec`, containing the values from 1 to 7.

```{r ex2a, echo=show_solution, eval=show_solution}
num_vec <- c(1, 2, 3, 4, 5, 6, 7)
```

b. Genarate a character vector, named `char_vec` with the days of the week.

```{r ex2b, echo=show_solution, eval=show_solution}
char_vec <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
```

c. Starting from the vector:  

    ```{r ex2c}
fac <- c("F", "F", "M", "M", "F", "F", "M")
    ```
  Generate the corresponding factor, named `fac`, with two levels: "F" and "M"

```{r ex2d, echo=show_solution, eval=show_solution}
fac <- factor(fac, levels = c("F", "M"))
fac
```

d. Generate a data frame, named `df2`, containing the previously defined: `num_vec`, `char_vec` and `fac`. Remember to maintain character vectors as they are, specifiyng `stringsAsFactors = FALSE`. 

```{r ex2e, echo=show_solution, eval=show_solution}
df <- data.frame(num_vec, char_vec, fac, stringsAsFactors = FALSE)
df
```

e. Supposing `dplyr` package is already installed and loaded, convert the previously defined data frame in tbl_df class. 

```{r ex2f, echo=show_solution, eval=show_solution}
df <- df %>% tbl_df()
df
```

## Matrices

### Exercise 1 

Generate a matrix, named `mat`, with 5 rows and 3 columns containing numbers from 1 to 15, using `matrix()` function.

```{r ex3a, echo=show_solution, eval=show_solution}
mat <- matrix(1:15, nrow = 5, ncol = 3, byrow = TRUE)
mat
```

## Lists

### Exercise 1 

Generate a list, named `my_list` that contains the following R elements:

```{r ex4a}
char <- "Veronica"
mat <- matrix(1:9, ncol = 3)
log_vec <- c(TRUE, FALSE, TRUE, TRUE)
```

```{r ex4b, echo=show_solution, eval=show_solution}
my_list <- list(char <- "Veronica", 
              mat = matrix(1:9, ncol = 3), 
              log_vec = c(TRUE, FALSE, TRUE, TRUE))
my_list
```




