
# Statistical models

```{r solution, echo=FALSE}
show_solution <- show_solution <- if (exists("show_solution_mk")){
  show_solution_mk
} else {
  FALSE
}
```

Before starting the exercises, load the following libraries, supposing they are already installed.

```{r require, message=FALSE}
require(dplyr)
require(ggplot2)
require(qdata)
```

## Linear Models

### Exercise 1

The number of impurities (lumps) present in the containers of paint depends on the rate of agitation applied to the container. A researcher wants to determine the relation between the rate of agitation and the number of lumps, so he conducts an experiment. He applies different rates of agitation (`Stirrate`) to 12 containers of paint and he counts the number of impurities (lumps) present in the containers of paint (`Impurity`).  

```{r data_import_lm1}
data(paint)
head(paint)
```

a. Let us compute the main descriptive statistics of `Impurity`. 

```{r desc_stats_lm1, echo=show_solution, eval=show_solution}
# Descriptive Statistics
summary_stat <- paint %>% summarise(n=n(),
  min = min(Impurity), 
  first_qu = quantile(Impurity, 0.25),
  median = median(Impurity),
  mean = mean(Impurity),
  third_qu = quantile(Impurity, 0.75),
  max = max(Impurity),
  sd = sd(Impurity))

print(summary_stat)
```

b. Let us graphically represent the relation between `Impurity` and `Stirrate` variables (add regression line to the scatterplot).

```{r reg_plot_lm1, echo=show_solution, eval=show_solution}
pl <- ggplot(data = paint, mapping = aes(x = Stirrate, y=Impurity)) +
  geom_point(color="blue") + 
  geom_smooth(method = "lm", colour="red", se = FALSE)
  
print(pl)  
# A simple straight linear regression is a good choice to describe the relation
```

c. Let us compute a simple linear regression between `Impurity` and `Stirrate`. 

```{r fit_lm1, echo=show_solution, eval=show_solution}
# Fit the linear model
fm <- lm(formula = Impurity ~ Stirrate, data = paint)
```

d. Does `Stirrate` influence `Impurity`? How? Let us analyze the model fitted by using `summary()` function.

```{r data_analysis_lm11, echo=show_solution, eval=show_solution}
summary(fm)
```


e. Let us check (final) models residuals.

```{r data_analysis_lm12, echo=show_solution, eval=show_solution}
# Residuals analysis
op <- par(mfrow = c(2,2))
plot(fm)
par(op)
```

### Exercise 2

A pressure switch has a membrane whose thickness (in mm) influences the pressure required to trigger the switch itself. The aim is to determine the thickness of the membrane for which the switch "trig" with a pressure equal to 165 ± 15 KPa. 25 switches with different thickness (`DThickness`) of the membrane was analysed, measuring the pressure at which each switch opens (KPa) (`SetPoint`).

```{r data_import_lm2}
data(switcht)
head(switcht)
```

a. Let us compute the descriptive statistics of `SetPoint` variable.

```{r desc_stats_lm2, echo=show_solution, eval=show_solution}
# Descriptive Statistics
summary_stat <- switcht %>% summarise(n=n(),
  min = min(SetPoint), 
  first_qu = quantile(SetPoint, 0.25),
  median = median(SetPoint),
  mean = mean(SetPoint),
  third_qu = quantile(SetPoint, 0.75),
  max = max(SetPoint),
  sd = sd(SetPoint))

print(summary_stat)
```

b. Let us graphically represent the relation between `DThickness` and `SetPoint`(add regression line to the graph).

```{r reg_plot_lm2, echo=show_solution, eval=show_solution}
pl <- ggplot(data = switcht, mapping = aes(x = DThickness, y=SetPoint)) +
  geom_point(color="blue") + 
  geom_smooth(method = "lm", colour="red", se = FALSE)
  
print(pl)  
```

c. Let us compute a linear regression between `DThickness` and `SetPoint` and check the residuals of the fitted model. 

```{r fit_lm2, echo=show_solution, eval=show_solution}
# Fit the linear model
fm1 <- lm(formula = SetPoint ~ DThickness, data = switcht)
```

d. Does `DThickness` influences `SetPoint`? Let us analyze the model fitted by using `summary()` function.

```{r data_analysis_lm21, echo=show_solution, eval=show_solution}
summary(fm1)
```

e. Let us check (final) models residuals.

```{r data_analysis_lm22, echo=show_solution, eval=show_solution}
# Residuals analysis
op <- par(mfrow = c(2,2))
plot(fm1)
par(op)
```

