---
title: "Tibbles"
---

```{r options, include=FALSE, purl=FALSE}
options(width = 108)
knitr::opts_chunk$set(message = FALSE)
```


# Introduction

Instead of the traditional R's data frames, throughout this course you will work mainly with "tibbles". "Tibbles" are similar to data frames but offer better checking and printing capabilities than traditional data frames. Almost all of the functions that you will use in this course produce, and take as input, tibbles. In fact tibbles are one of the unifying features of the tidyverse. 

In this chapter you will work with the `tibble` package which provides functions for tibbles coercion and creation.

You can load `tibble` by loading `tidyverse`:

```{r}
require(tidyverse)
```

or you can load it explicitely, together with `dplyr` that will be used throughout this chapter:

```{r}
require(tibble)
require(dplyr)
```


# `tbl_df`: the `tidyverse` Data Frame Class

`dplyr`, as well as other packages that you will work with in this course, can work with data frames as is but it is worthwhile to convert them to a tibble (`tbl_df`), especially if frames have large dimensions. `tbl_df` is a wrapper around a data frame that, as mentioned, will not accidentally print a lot of data to the screen, indeed tbl objects only print a few rows and all the columns that fit in one screen, describing the rest of it as text.


When the class of data object is not tibble, you can convert it to a tibble with the function `as_tibble()`.  
Let us consider `mtcars`, a dataset included in `datasets` package (automatically loaded at the start of an R session): 

```{r data frame, message=FALSE}
# Example of data frame
class(mtcars)

# If we do not convert it as a tbl_df, all mtcars rows and columns will be printed when calling mtcars 
dim(mtcars)
mtcars
```

Now, using the package `tibble`, you can convert mtcars into a `tbl_df` object:

```{r tibble, message=FALSE}
# tibble version of the same data frame 
mtcars_tbl <- as_tibble(mtcars)
mtcars_tbl
```

Alternatively, you may use the function `tbl_df()`  in the package `dplyr`.

```{r tbl_df, message=FALSE}
# tibble version of the data frame mtcars with the function tbl_df()
mtcars_tbl2 <- tbl_df(mtcars)
class(mtcars_tbl)
mtcars_tbl
```

Moreover you may want to create a tibble from individual vectors. This can be easily done with the function `tibble()`.

```{r tibble2, message=FALSE}
# tibble version of the data frame mtcars with the function tbl_df()
tb <- tibble(
  id = 1:5,
  height = c(1.7, 1.65, 1.6, 1.75, 1.73),
  weight = c(80, 46, 52, 82, 75),
  bmi = weight/ (height) ^2
)
tb
```

Notice that it is possibile for a tibble to have column names that are not valid as `R` variables names, for instance they may be numbers or may contain unsual characters such as spaces.

```{r tibble3, message=FALSE}
# tibble version of the data frame mtcars with the function tbl_df()
tb <- tibble(
  ":)" = "happy",
  "1" = "one",
  "/" = "slash",
  " " = "space"
)
tb
```
