---
title: "Data Management with `dplyr` and `tidyr`"
---

```{r options, include=FALSE, purl=FALSE}
options(width = 108)
```


# Introduction

This course will provide an overview of data management with `R` through the `dplyr` and `tidyr` packages.  

`dplyr` is a modern `R` package which provides functions to manage data in a more intuitive and efficient way than base `R`.  
The starting point for more information about `dplyr` is the official website providing documentation: [dplyr: A Grammar of Data Manipulation](https://cran.r-project.org/web/packages/dplyr/index.html).  

`tidyr` is a modern `R` package which provides a standard way to organise data values within a dataset.  
The starting point for more information about `tidyr` is the official website providing documentation: [tidyr: Easily Tidy Data with 'spread()' and 'gather()' Functions](https://cran.r-project.org/web/packages/tidyr/index.html).  

Supposing the packages are already installed, first of all `dplyr` and `tidyr` must be loaded.

```{r data, message=FALSE}
require(dplyr)
require(tidyr)
```


# Datasets Used in the Examples of This Course

The following data sets are used in the examples of this course.

## The `bank` Dataset

Most of the examples of this course require the dataset `bank` which is included in the `qdata` package.

```{r bank, message=FALSE}
require(qdata)
data(bank)
bank
str(bank)
```

This dataset contains information on phone calls related with direct marketing campaigns of a Portuguese banking institution. The data frame has 20 variables and 45211 observations. The marketing campaigns were based on phone calls. The data are ordered by call id (and also by date, from May 2008 to November 2010). The full dataset was described and analyzed in:

S. Moro, R. Laureano and P. Cortez. Using Data Mining for Bank Direct Marketing: An Application of the CRISP-DM Methodology. In P. Novais et al. (Eds.), Proceedings of the European Simulation and Modelling Conference - ESM'2011, pp. 117-121, Guimarães, Portugal, October, 2011. EUROSIS.


## Other Datasets

Some single sections of this course contain examples also refered to other datasets included in the `qdata` package:

* `people`, which is related with information on gender, area, weight and height of 1806 Italian people:

```{r people, message=FALSE}
data(people)
people
str(people)
```

* `mtcars`, which was extracted from the analogous data frame of package `datasets`; it contains information on 32 cars:

```{r mtcars, message=FALSE}
data(mtcars)
mtcars
str(mtcars)
```

* `comuni`, `province`, `regioni` and `ripartizioni`, which contain information on the Italian districts (8003 "comuni", 110 "province", 20 "regioni" and 5 "ripartizioni"). They were all obtained as subsets or by summarizing the `italia` data frame:

```{r italia, message=FALSE}
data(italia)
italia
str(italia)
```

* `usopen` and `wimbledon`, which contain information about the gentlemen U.S. Open nd Wimbledon finals from 1968 to 2015:

```{r tennis, message=FALSE}
data(usopen)
usopen
str(usopen)
data(wimbledon)
wimbledon
str(wimbledon)
```

For a complete description of these datasets please refer to the help:

```{r help, eval=FALSE}
?bank
?people
?mtcars
?italia
?comuni
?province
?regioni
?ripartizioni
?usopen
?wimbledon
```
