---
title: "Conditionals"
---

```{r options, include=FALSE, purl=FALSE}
options(width = 108)
```

## The `if` statement
Conditionals are very useful when programming, as they allow to execute some code when a condition is satisfied.

The `if` statement is the most used conditionals. Its use is quite simple:

```{r if0, eval = FALSE}
if(condition) code to be executed
```

The condition ought be a lenght-one logical vector. Conditions of length greater than one are accepted with a warning, but only the first element is used. If condition is `TRUE` then the code is executed.
```{r if}
x <- 2
if (x > 0) print("x greater than zero")
```

The code above print the text, because the condition is satisfied. When the condition is false, the code is ignored.
```{r iffalse}
x <- -4
if (x > 0) print("x greater than zero")
```

When condition is longer than a one line of code, the code to be executed ought be enclosed in braces (`{...}`). 
```{r ifline}
x <- 2
if (x > 0) {
  print("x greater than zero")
  return(x)
}
```

The `if` statement allows an `else` condition that will be executed when the condition is false.

```{r else0, eval = FALSE}
if(condition) code-when-true else code-when-false
```


As an example:
```{r ifelse}
x = -4
if (x > 0) print("x greater than zero") else print("x is not greater than zero")
```

In this case, it is strongly suggested to always use braces.
```{r ifelsebraces}
x <- -4
if (x > 0) {
  print("x greater than zero")
} else {
  print("x is not greater than zero")
}
```

Of course, the `if` statement can be used to assign values to an object.
```{r ifassignement}
x <- -4
if (x > 0) {
  y <- 1
} else {
  y <- -1
}
y
```

## The `ifelse()` function

As seen above, the `if` statement applies only to a length-one logical vector. If applied to vectors with length greater than one, only the first element is used and a warning message is returned. 

```{r structures-008}
if(x < 5){"LESS"} else {"GREATER"}
```

The `ifelse()` function allows us to deal with vectors of any length. Its arguments are: a test condition, a return value when the test returns TRUE, a return value when the test returns FALSE.

```{r structures-009}
x = 1:10
ifelse(x < 5, "LESS", "GREATER")
```

Moreover, the `ifelse()` function return a vector, so it can be assigned to an object.
```{r ifelsefunction}
x <- -4
y <- ifelse(x > 0, 1, -1)
y
```

## Select one of a list of alternatives: `switch`

The `switch()` function evaluates an expression and accordingly chooses one of the further arguments. It is usually used inside a function to select from a list of alternatives.

The following code, provide a self-written `centre()` function, that accepts two input arguments: a numeric vector and a centroid function  (mean, median, trimmed mean) to be applied to the numeric vector.

The `switch` read the string with the centroid function (`type`) and applies the accordingly function.

```{r switch, tidy=FALSE}
centre <- function(x, type) {
  switch(type,
    mean = mean(x),
    median = median(x),
    trimmed = mean(x, trim = .1)
  )
}
```

```{r centre, echo=2:100}
set.seed(355)
x <- rcauchy(10)
centre(x, "mean")
centre(x, "median")
centre(x, "trimmed")
```