---
title: "summarise_all and mutate_all"
---

```{r options, include=FALSE, purl=FALSE}
options(width = 108)
```

```{r first, include=TRUE, purl=TRUE, message=FALSE}
require(dplyr)
```

In addition to `summarise_at()` and `mutate_at()` two other functions were added `summarise_all()` and `mutate_all()`. They apply a set of desired functions on all the variables in the dataset.

## `summarise_all()` and `mutate_all()`

Consider the following data:

```{r}
my_data <- data.frame(id = 1:10, 
                      gender = c(rep("F",5), rep("M", 5)),
                      x = rbinom(10, 1, 0.5), 
                      y = rnorm(10, 5, 1))

my_data <- tbl_df(my_data)
```

Suppose we want to find the mean of all the variables:

```{r}
my_data %>% 
  summarise_all(mean) # warning for non numeric

my_data %>% 
  select(c(x, y)) %>%
  summarise_all( funs(mean, sd)) # you do not need to specify names for the variables
```

Similarly, suppose you want to apply a specific mutation to all the variables in the dataset:

```{r}
squared <- function(x) {x^2}

my_data %>% 
  select(c(x, y)) %>%
  mutate_all(funs(squared))
```
