---
title: "Your first R session"
---

```{r setup, include=FALSE, purl=FALSE, message=FALSE}
require(knitr)
opts_chunk$set(echo = TRUE)
```


## Aritmetic with R

Start the R system, the cursor is waiting for you to type in some R commands. For example, use R as a simple calculator:

```{r calculator}
6 + 3
5 - 9
4 * 6
8 / 3
5 ^ 2
(1 + 0.05)^8
exp(3)
log(14)
23.76 * log(8)/(23 + atan(9))
```

## Assignment

Results of calculations can be stored in objects using the assignment operator:

```{r assignment}
x <- log(14)
y <- 23.76 * log(8)/(23 + atan(9))
```

To print the object just enter the name of the object or use `print()` function. 

```{r print_obj}
x
print(y)
```

These objects can then be used in other calculations. 

```{r calc_obj}
z <- x + y
z
```

## The R Workspace

The workspace is your current R working environment and includes any user-defined objects. It is also known as global environment.

### Objects listing

Objects created during an R session are hold in memory. To list the objects in the current R session, the function `ls()` or the function `objects()` may be used.

```{r workspace}
ls()
objects()
```


### Removing objects

If a value to an object that already exists is assigned then the contents of the object will be overwritten with the new value (without a warning!). The function `rm()` ought be used to remove one or more objects from your session.

```{r rm}
rm(x)
ls()
```


## R help

Within R, the following functions provide help about R itself:

 - The HTML version of R's online documentation can be printed on-screen by typing `help.start()`;
 - Online documentation for most of the functions and variables in R can be printed on-screen by typing `help(name)` (or `?name`), where _name_ is the name of the topic help is sought for;
 - Online documentation for finding help pages on a vague topic can be printed on-screen by typing `help.search('topic')`;
 - A list of function containing _topic_ in the name can be printed on-screen by typing `apropos('topic')`;
 - A research in the website can be performed by typing `RSiteSearch('query')`, where _query_ is the search query.

For example, to get help about the `mean()` function, the `help()` function can be used.

```{r help1, eval=FALSE}
help(mean)
```

The `help()` function can be called using the `?`.
```{r help2, eval=FALSE}
?mean
```

To get a list of functions concerning the mean, the `help.search()` function can be used.

```{r help.search1, eval=FALSE}
help.search("mean")
```

To get a list of function containing "mean" in the name, the `apropos()` function can be used.

```{r help.search2, eval=FALSE}
apropos("mean")
```
