---
title: "null"
author: "null"
date: "null"
output:
  pdf_document: default
  html_document:
    self_contained: no
---

```{r setup, echo=FALSE, message=FALSE, results='hide'}
require(knitr) 
options(width=80)
opts_chunk$set(list(dev = 'png', fig.cap='', fig.show='hold', dpi=100, fig.width=3, fig.height=2.5, fig.pos='H!'))#, fig.path="figures/lm-"))
show_solution <- TRUE
```
 
# Writing R functions
## Exercise 1

Write a function, named `compute_summary`, which computes: sum, subtraction, multiplication and division of two numbers. The function arguments should be the two numbers, named as: `x` and `y`. The function should return all amounts computed. 

```{r ex1, echo=show_solution, eval=show_solution}
compute_summary <- function(x, y){
  sum_op <- x+y
  sub_op <- x-y
  mul_op <- x*y
  div_op <- x/y
  return(list(sum_op=sum_op, sub_op=sub_op, mul_op=mul_op, div_op=div_op))
}

compute_summary(x=4, y=2)
compute_summary(x=3, y=7)
```

## Exercise 2

Write a function, named `compute_gain`, which computes the income by multiplying the amount produced for sale price and then computes the gain by subtracting the costs to income.  
The function arguments should be: `amount`, `price`, and `costs`; `price` should have a default value equal to 5. The function should return the gain.

```{r ex2, echo=show_solution, eval=show_solution}
compute_gain <- function(amount, costs, price=5){
  income = amount * price
  gain = income - costs
  return(gain)  
}

compute_gain(amount = 40, costs = 50)
compute_gain(amount = 100,costs = 70,price = 1)

```

